Efficient implementation of kernel filtering

### 2007-07-02 article/Article.tex
add a few more graphics
darcs-hash:20070702214843-2fc9d-ee884ff04584fda5be296a972d63adcbfc66a12b.gz
minor grammar fixes to article
darcs-hash:20070702235032-fafb9-6a51baa1460da340327dbd7ab3353889b2fb2d30.gz

% Increment the release number whenever significant changes are made.
% The author and/or editor can define 'significant' however they like.
\release{0.00}

@@

\section{Introduction}
A kernel based filtering process replaces the pixel at the kernel
origin with the result of applying a function to all pixels defined by
the kernel. Many useful filters, including edge detection and gradient
filters, smoothing filters and rank and morphology filters fall into
this category. Direct implementations of such filters typically
involve visiting all pixels defined by the kernel in order to evaluate
the filter function. Such an approach is usually easy to implement
-- in ITK it is made simple by the neighborhood iterators -- but leads
to an algorithm complexity proportional to the number of pixels in the
kernel (or $O(n^d)$, where $n$ is the kernel size and $d$ is the
dimensionality). Such complexity tends to restrict application of such
filters to small kernels.

@@

\subsection{Arbitrary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an up-datable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array, with place holders
tracking the current maximum or minimum increasing performance. More
sophisticated histogram representations are necessary for larger pixel
types. Our implementation uses c++ maps, which is an extension to the
original paper.

@@

Approximations of more complex shapes, particularly circles and
ellipses can constructed using a number of lines at evenly spaced
angles \cite{Adams93}. It is difficult to create a structuring element
with a precisely defined radius using this method because line
structuring elements from which the circle structuring element is
composed must have odd length and there are practical limits due to
the realities of underlying digital grid representation of images. In
addition it is possible that the structuring elements may not be truly
translation invariant due to the representation of line
(e.g. Bresenham) used in the decomposition. However, precisely defined
radii are rarely critical when a large structuring element is called
for. An example of a structuring element created using line
structuring elements is shown in Figure
\ref{fig:circledecomposition}. Composition of regular shapes, such as
hexagons and octagons is more accurate.

@@

@@

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{accum2D}
\caption{Recursive computation of the accumulator image - value at $(i,j)$ computed using Equation \ref{eq:accum}. The double hashed region is subtracted twice because it is included in both $b_{i,j-1}$ and $b_{i-1, j}$.\label{fig:accum}}
\end{figure}

@@

\section{ITK implementation}
The filters discusses in this article are contained in 3 packages --
{\em consolidatedMorphology}, {\em fastRankMean} and {\em
boxConvolution}. The most recent versions of these packages can be
obtained from \url{http://voxel.jouy.inra.fr/darcsweb/}\footnote{
The most recent versions can be obtained using darcs \cite{DarcsWebSite}
with the command {\em darcs get http://voxel.jouy.inra.fr/\\darcs/contrib-itk/package},
where {\em package} must be replace by {\em consolidatedMorphology},
{\em fastRankMean} or {\em boxConvolution}.}.
All of the packages
include multithreaded filters and include many examples and tests. Unless
indicated otherwise, all filters will behave basically the same way as
other kernel filters, requiring either a radius or a neighborhood to
be specified.

\subsection{Morphology filters -- consolidatedMorphology}
An earlier version of the {\em consolidatedMorphology} package has
been submitted to the InsightJournal. However many improvements and
bug fixes have been included since then.

The morphology filters utilize a new class -- {\em
FlatStructuringElement} to describe, arbitary and decomposable
structuring elements. It provides the following methods to create
structuring elements with different characteristics:
\begin{itemize}
  \item \verb$Ball()$ generate a ball structuring element. It takes the 
	radius of the structuring element as parameter.
  \item \verb$Box()$ produces a box structuring element. It takes the radius 
	of the structuring element as parameter.
  \item \verb$Poly()$ produces an approximation to a circle or sphere using 
	line structuring elements, taking the desired radius and number of 
	lines as parameters. In the 3D case an approximation of a sphere will 
	be attempted. Caution is advised because this is not well tested yet.
  \item \verb$FromImage()$ produces a structuring element from an image. The image is passed
as parameter to the method. A optional parameter can also be passed: the pixel value
to be considered as the foreground value in the image. This value defaults to
\verb$NumericTraits< PixelType >::max()$. This lets the user visualize the structuring element.
\end{itemize}

The algorithm used by a filter is set by the {\em SetAlgorithm}
method, with options of {\em BASIC}, {\em HISTO}, {\em VHGW} and {\em
ANCHOR} for the traditional ITK implementation, the moving histogram
algorithm, the van Herk, Gil, Werman algorithm and the anchor
algorithm respectively. The VHGW and ANCHOR algorithms can only be
used for Box and Poly structuring elements. VHGW is more thoroughly
tested. This HISTO algorithm produces the identical results to BASIC,
and is always faster. It is the default and should be used if you need
precisely defined, non rectangular, structuring elements. If a Box
structuring element is needed then VHGW or ANCHOR should be used, as
they offer the best performance and are exact in that case. Poly
structuring elements are approximations of circles that will offer
improved performance with VHGW or ANCHOR algorithms, but the user will
need to consider whether the approximation is suitable for the
application in question.

@@

\subsection{Morpholgy}
\begin{table}[htbp]
\centering
\begin{tabular}{cccc}
\hline
Radius & Histogram & vanHerk/Gil-Weman & Anchor \\
1    &   0.0295  & 0.0228  & 0.033 \\
2    &   0.0219  & 0.0126  & 0.0187 \\
3    &   0.028   & 0.0128  & 0.0167 \\
4    &   0.0339  & 0.0262  & 0.0379 \\ 
5    &   0.0403  & 0.0264  & 0.0363 \\
6    &   0.046   & 0.0265  & 0.038 \\
7    &   0.052   & 0.027   & 0.0343 \\
8    &   0.0637  & 0.0275  & 0.0344 \\
9    &   0.0665  & 0.0387  & 0.0599 \\
10   &   0.117   & 0.0641  & 0.142 \\
15   &   0.103   & 0.0385  & 0.0455 \\
20   &   0.133   & 0.0409  & 0.0445 \\
25   &   0.164   & 0.0399  & 0.044 \\
30   &   0.199   & 0.0403  & 0.0427 \\
40   &   0.263   & 0.0405  & 0.0425 \\
50   &   0.329   & 0.0403  & 0.043 \\
100  &   0.68    & 0.0404  &  0.0472 \\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for dilation by circles using histogram, and polygon approximation based implementations applied to the $256 \times 256$ cthead image on a 1.8GHz AMD Athlon, 512KB cache. The polygon approximations use the default number of lines (a maximum of 6 when radius is greater than 8.\label{tab:perfPoly}}
\end{table}

\begin{figure}[htbp]
\begin{center}
\subfigure[Dilation]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-dilation-small}}
\subfigure[Opening]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-opening-small}}
\subfigure[Gradient]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-gradient-small}}
\subfigure[TODO]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-gradient-small}}
\subfigure[TODO]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-gradient-small}}
\subfigure[TODO]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-gradient-small}}
\caption{TODO}
\end{center}
\end{figure}

\subsection{Rank and Mean filters}
Relative times for sliding window median and mean filters and
separable median filters are shown in Table \ref{tab:perfRank}. The
predicted linear complexity is observed for the sliding window
approaches (complexity of direct approach is $O(n^2)$ in 2D, reduced
to $O(n)$ by using the sliding window. The separable median exhibits a
runtime independent of kernel size, as expected.
\begin{table}[htbp]
\centering
\begin{tabular}{cccccc}
\hline
Size  &  Direct Median &  Sliding Median & Separable Median & Direct Mean & Sliding Mean\\
1     &   0.0352 & 0.0341 & 0.0378 & 0.0171 & 0.0229\\
5     &   0.409  & 0.0823  & 0.0201 & 0.24  &  0.0601 \\
10    &   1.03   & 0.131 & 0.0193 & 0.803 &  0.0569 \\
15    &   2.82   & 0.106  & 0.0192 & 2.13  &  0.0819 \\
20    &   5.85   & 0.138  & 0.0358 & 4.76  &  0.106 \\
40    &   39.7   & 0.2563  & 0.0198 & 36.6  &  0.203 \\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for median and mean filters using direct and sliding window implementations applied to the $256 \times 256$ cthead image on a 1.8GHz AMD Athlon, 512KB cache. The separable median is an approximation and therefore doesn't produce the same result as the direct or sliding algorithms.\label{tab:perfRank}}
\end{table}

\subsection{Box Convolution}
The relative times for optimized and standard convolution are shown in
Table \ref{tab:perfBoxConv}, and demonstrate significantly improved
performance for kernel radii greater than 1. The times support the
theoretical prediction that the complexity of the algorithm is
independent of kernel size.
\begin{table}[htbp]
\centering
\begin{tabular}{ccccc}
\hline
Size  &  Direct Mean &  Box Mean & Direct sigma & Box sigma\\
1    &   0.0142 & 0.0226 & 0.0176 & 0.0332 \\
2    &   0.0279 & 0.0176 & 0.0294 & 0.0258 \\
3    &   0.0425 & 0.013  & 0.0447 & 0.0193 \\
4    &   0.077  & 0.0132 & 0.0803 & 0.0195 \\
5    &   0.128  & 0.0132 & 0.129  & 0.0195 \\
6    &   0.195  & 0.0134 & 0.195  & 0.0198 \\
7    &   0.275  & 0.0136 & 0.283  & 0.0198 \\
8    &   0.392  & 0.0148 & 0.405  & 0.0204 \\
9    &   0.54   & 0.014  & 0.548  & 0.0201 \\
10   &   0.706  & 0.0139 & 0.713  & 0.0202 \\
15   &   2.12   & 0.0158 & 2.18   & 0.021 \\
20   &   4.62   & 0.0157 & 4.76   & 0.0217 \\
25   &   8.63   & 0.0164 & 8.91   & 0.0222 \\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for mean and standard deviation using direct and box convolution implementations applied to the $256 \times 256$ cthead image on a 1.8GHz AMD Athlon, 512KB cache.\label{tab:perfBoxConv}}
\end{table}
\subsection{Timing notes}
You may have noticed that the optimized filters of radius 1 seem to
regularly perform slower than their larger counterparts. We can think
of no reason for this. The algorithms that have kernel size
independent complexity should not be slower for small kernels than
large ones because the only difference occurs at borders and when the
sliding window is being created, and both situations favour small
kernels. We can only guess that there are caching issues involved.

@@

\section{Acknowledgments}
We thank Dr Pierre Adenot and MIMA2 confocal facilities
(\url{http://mima2.jouy.inra.fr}) for providing the 3D test image.
We are grateful to the INRA MIGALE bioinformatics platform
(\url{http://migale.jouy.inra.fr}) for providing the computational resources
used for the timing tests.
\appendix

>>>>

% Increment the release number whenever significant changes are made.
% The author and/or editor can define 'significant' however they like.
% \release{0.00}

@@

\section{Introduction}
A kernel based filtering process replaces the pixel at the kernel
origin with the result of a function of all pixels defined by
the kernel. Many useful filters, including edge detection and gradient
filters, smoothing filters and rank and morphology filters fall into
this category. Direct implementations of such filters typically
involve visiting all pixels defined by the kernel in order to evaluate
the filter function. Such an approach is usually easy to implement
-- in ITK it is made simple by the neighborhood iterators -- but leads
to an algorithm complexity proportional to the number of pixels in the
kernel (or $O(n^d)$, where $n$ is the kernel size and $d$ is the
dimensionality). Such complexity tends to restrict application of such
filters to small kernels.

@@

\subsection{Arbitrary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an up-datable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array and the histogram updated
by incrementing or decrementing the appropriate bins. Using place
holders to track the current maximum or minimum increases
performance. More sophisticated histogram representations are
necessary for larger pixel types. Our implementation uses c++ maps,
which is an extension to the original paper.

@@

Approximations of more complex shapes, notably circles and
ellipses can constructed using a number of lines at evenly spaced
angles \cite{Adams93}. It is difficult to create a structuring element
with a precisely defined radius using this method because line
structuring elements from which the circle structuring element is
composed must have odd length and there are practical limits due to
the realities of underlying digital grid representation of images. In
addition it is possible that the structuring elements may not be truly
translation invariant due to the representation of line
(e.g. Bresenham) used in the decomposition. However, precisely defined
radii are rarely critical when a large structuring element is called
for. An example of a structuring element created using line
structuring elements is shown in Figure
\ref{fig:circledecomposition}. Composition of regular shapes, such as
hexagons and octagons is more accurate.

@@

\begin{figure}[htbp]
\begin{center}
\subfigure[Input]{\includegraphics[scale=1.88]{CellsFluorescence2}}
\subfigure[Dilation]{\includegraphics[scale=0.45]{dilate2D-std}}
\subfigure[Erosion]{\includegraphics[scale=0.45]{erode2D-std}}
\subfigure[Opening]{\includegraphics[scale=0.45]{open2D-std}}
\subfigure[Closing]{\includegraphics[scale=0.45]{close2D-std}}
\subfigure[White top-hat]{\includegraphics[scale=0.45]{whiteTopHat2D-std}}
% \subfigure[Gradient]{\includegraphics[scale=0.5]{gradient2D-std}}
\caption{Effect of a dilation (b), an erosion (c), an opening (d),
a closing (e) and a white top hat (f) with a ball structuring element.}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{accum2D}
\caption{Recursive computation of the accumulator image - value at $(i,j)$ computed using Equation \ref{eq:accum}. The double hashed region is subtracted because it is included in both $b_{i,j-1}$ and $b_{i-1, j}$.\label{fig:accum}}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\subfigure[Input]{\includegraphics[scale=0.6]{cthead1}}
\subfigure[Mean]{\includegraphics[scale=0.6]{test}}
\subfigure[Sigma]{\includegraphics[scale=0.6]{sigma}}
\caption{A mean transform (b) and a standard deviation transform (c) with a kernel of radius 3.}
\end{center}
\end{figure}

\section{ITK implementation}
The filters discusses in this article are contained in 3 packages --
{\em consolidatedMorphology}\footnote{An earlier version of the {\em consolidatedMorphology} package has
been submitted to the InsightJournal. However many improvements and
bug fixes have been included since then.}, {\em fastRankMean} and {\em
boxConvolution}. The most recent versions of these packages can be
obtained from \url{http://voxel.jouy.inra.fr/darcsweb/}\footnote{
The most recent versions can be obtained using darcs \cite{DarcsWebSite}
with the command {\em darcs get http://voxel.jouy.inra.fr/\\darcs/contrib-itk/package},
where {\em package} must be replace by {\em consolidatedMorphology},
{\em fastRankMean} or {\em boxConvolution}.}.
All of the packages
include multithreaded filters and include many examples and tests. Unless
indicated otherwise, all filters will behave basically the same way as
other kernel filters, requiring either a radius or a neighborhood to
be specified.

\subsection{Morphology filters -- consolidatedMorphology}

The morphology filters utilize a new class -- {\em
FlatStructuringElement} to describe, arbitary and decomposable
structuring elements. It provides the following methods to create
structuring elements with different characteristics:
\begin{itemize}
  \item \verb$Ball()$ generate a ball structuring element (a circle in 2D). It takes the 
	radius of the structuring element as parameter.
  \item \verb$Box()$ produces a box structuring element. It takes the radius 
	of the structuring element as parameter.
  \item \verb$Poly()$ produces an approximation to a circle or sphere using 
	line structuring elements, taking the desired radius and number of 
	lines as parameters. In the 3D case an approximation of a sphere will 
	be attempted. Caution is advised because this is not well tested yet.
  \item \verb$FromImage()$ produces a structuring element from an image. The image is passed
as parameter to the method. A optional parameter can also be passed: the pixel value
to be considered as the foreground value in the image. This value defaults to
\verb$NumericTraits< PixelType >::max()$. This lets the user visualize the structuring element.
\end{itemize}

The algorithm used by a filter is set by the {\em SetAlgorithm}
method, with options of {\em BASIC}, {\em HISTO}, {\em VHGW} and {\em
ANCHOR}, for the traditional ITK implementation, the moving histogram
algorithm, the van Herk, Gil, Werman algorithm and the anchor
algorithm respectively. The VHGW and ANCHOR algorithms can only be
used for Box and Poly structuring elements. VHGW is more thoroughly
tested. The HISTO algorithm produces the identical results to BASIC,
and is always faster. It is the default and should be used if you need
precisely defined, non rectangular, structuring elements. If a Box
structuring element is needed then VHGW or ANCHOR should be used, as
they offer the best performance and are exact in that case. Poly
structuring elements are approximations of circles that will offer
improved performance with VHGW or ANCHOR algorithms, but the user will
need to consider whether the approximation is suitable for the
application in question.

@@

\subsection{Morpholgy}

Execution times for the dilation, the opening, and the gradient transform are shown in
Tables \ref{tab:perfDilation}, \ref{tab:perfOpening} and \ref{tab:perfGradient}, and in
Figure \ref{fig:consolidatedMorphology-perf}. The predicted linear complexity for the number
of neighbors is observed for the basic algorithm, as well as the constant complexity of the
van Herk / Gil Werman and the anchor algorithm. The linear complexity for the number of
pixels added and removed per translation is observed for the moving histogram algorithm,
as expected.

A good threading support is observed for all the implementations, as shown in Table
\ref{tab:perfThreads} and Figure \ref{fig:consolidatedMorphology-perf}.

\begin{figure}[htbp]
\begin{center}
\subfigure[Dilation]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-dilation-small}}
\subfigure[Opening]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-opening-small}}
\subfigure[Gradient]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2D-gradient-small}}
\subfigure[Threads]{\includegraphics[scale=0.62]{consolidatedMorphology-perf_threads}}
\subfigure[Number of neighbors]{\includegraphics[scale=0.62]{consolidatedMorphology-perf_nb_of_neighbors}}
\subfigure[Polygon]{\includegraphics[scale=0.62]{consolidatedMorphology-perf2Dpoly}}
\caption{Execution times with a box structuring element of increasing size for different algorithms of dilation (a),
opening (b) and gradient (c).
Execution times with and increasing number fo threads (d).
Execution times with an increasing number of neighbors in a box structuring of constant radius (e).
Execution times with a polygon structuring element. The polygon approximations use the default number of lines (a maximum of 6 when radius is greater than 8 (f).\label{fig:consolidatedMorphology-perf}
Results are reported in Tables \ref{tab:perfDilation}, \ref{tab:perfOpening}, \ref{tab:perfGradient},
\ref{tab:perfThreads} and \ref{tab:perf2Dpoly}.
}
\end{center}
\end{figure}

\subsection{Rank and Mean filters}
Relative times for sliding window median and mean filters and
separable median filters are shown in Tables \ref{tab:perfRankMean}
and \ref{tab:perfRankMedian} and in Figure \ref{fig:perfRank}. The
predicted linear complexity is observed for the sliding window
approaches (complexity of direct approach is $O(n^2)$ in 2D, reduced
to $O(n)$ by using the sliding window. The separable median exhibits a
runtime independent of kernel size, as expected.

\begin{figure}[htbp]
\begin{center}
\subfigure[Mean]{\includegraphics[scale=0.62]{fastRankMean-perfMean}}
\subfigure[Median]{\includegraphics[scale=0.62]{fastRankMean-perfMedian}}
\caption{Execution times for several algorithms of mean filters (a) and
median filters (b).\label{fig:perfRank}
Results are reported in Tables \ref{tab:perfRankMean} and  \ref{tab:perfRankMedian}.}
\end{center}
\end{figure}

\subsection{Box Convolution}
The relative times for optimized and standard convolution are shown in
Table \ref{tab:perfBoxConv} and Figure \ref{fig:perfBoxConv}, and demonstrate significantly improved
performance for kernel radii greater than 1. The times support the
theoretical prediction that the complexity of the algorithm is
independent of kernel size.

A good threading support is observed for all the implementations, as shown in Table
\ref{tab:perfBoxConv-threads} and Figure \ref{fig:perfBoxConv}.

\begin{figure}[htbp]
\begin{center}
\subfigure[Neighborhood size]{\includegraphics[scale=0.62]{boxConvolution-perf2D-large}}
\subfigure[Threads]{\includegraphics[scale=0.62]{boxConvolution-perf_threads-large}}
\caption{Execution times for and increasing kernel size (a), and an increasing number of threads (b).
\label{fig:perfBoxConv}
Results are reported in Tables \ref{tab:perfBoxConv} and  \ref{tab:perfBoxConv-threads}}
\end{center}
\end{figure}

\subsection{Timing notes}
You may have noticed that the optimized filters of radius 1 seem to
regularly perform slower than their larger counterparts. We can think
of no reason for this. The algorithms that have kernel size
independent complexity should not be slower for small kernels than
large ones because the only difference occurs at borders and when the
sliding window is being created, and both situations favour small
kernels. We can only guess that there are caching issues involved.

@@

\section{Acknowledgments}
We thank Dr Pierre Adenot and MIMA2 confocal facilities
(\url{http://mima2.jouy.inra.fr}) for providing the 3D test image.
We are grateful to the INRA MIGALE bioinformatics platform
(\url{http://migale.jouy.inra.fr}) for providing the computational resources
used for the timing tests.

\appendix

\begin{table}[phtb]
\centering
\begin{tabular}{ccccccc}
\hline
Radius	&	Basic	&	Separable	&	Histogram	&	Separable histogram	&	Anchor	&	van Herk / Gil Werman	\\
\hline
1	&	0.00477	&	0.00491	&	0.013	&	0.012	&	0.0144	&	0.00725	\\
2	&	0.0114	&	0.0101	&	0.0161	&	0.0119	&	0.0135	&	0.00743	\\
3	&	0.024	&	0.0146	&	0.0193	&	0.0116	&	0.0129	&	0.00794	\\
4	&	0.0444	&	0.0192	&	0.0231	&	0.0114	&	0.0126	&	0.00755	\\
5	&	0.0714	&	0.0247	&	0.0261	&	0.0114	&	0.0121	&	0.0074	\\
6	&	0.107	&	0.0299	&	0.0313	&	0.0116	&	0.0127	&	0.00771	\\
7	&	0.153	&	0.034	&	0.0358	&	0.0113	&	0.0121	&	0.00859	\\
8	&	0.211	&	0.0407	&	0.038	&	0.0112	&	0.0121	&	0.00773	\\
9	&	0.281	&	0.0442	&	0.0434	&	0.0114	&	0.012	&	0.00779	\\
10	&	0.365	&	0.0486	&	0.0446	&	0.0111	&	0.0118	&	0.00763	\\
15	&	1.03	&	0.074	&	0.0604	&	0.0115	&	0.0116	&	0.00794	\\
20	&	2.23	&	0.104	&	0.079	&	0.0113	&	0.0113	&	0.00762	\\
25	&	4.11	&	0.129	&	0.0968	&	0.0114	&	0.0116	&	0.0079	\\
30	&	6.82	&	0.154	&	0.111	&	0.0114	&	0.0115	&	0.00781	\\
40	&	15.4	&	0.214	&	0.153	&	0.0116	&	0.0116	&	0.00785	\\
50	&	29.3	&	0.285	&	0.19	&	0.0119	&	0.0117	&	0.00804	\\
100	&	226	&	0.767	&	0.355	&	0.0129	&	0.0117	&	0.00809	\\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for dilation by boxes using several algorithm implementations applied to the $256 \times 256$ cthead image.\label{tab:perfDilation}}
\end{table}

\begin{table}[phtb]
\centering
\begin{tabular}{ccccc}
\hline
Radius	&	Basic	&	Histogram	&	Anchor	&	van Herk / Gil Werman	\\
\hline
1	&	0.0107	&	0.0124	&	0.0219	&	0.014	\\
2	&	0.0235	&	0.0167	&	0.0201	&	0.0144	\\
3	&	0.0488	&	0.0211	&	0.019	&	0.0154	\\
4	&	0.0895	&	0.0247	&	0.0184	&	0.0148	\\
5	&	0.143	&	0.0273	&	0.0176	&	0.0143	\\
6	&	0.216	&	0.037	&	0.0184	&	0.0149	\\
7	&	0.308	&	0.0372	&	0.0176	&	0.0167	\\
8	&	0.423	&	0.0401	&	0.0172	&	0.015	\\
9	&	0.563	&	0.0466	&	0.0171	&	0.0151	\\
10	&	0.73	&	0.0455	&	0.0166	&	0.0148	\\
15	&	2.07	&	0.0624	&	0.0165	&	0.0153	\\
20	&	4.46	&	0.0885	&	0.0159	&	0.0148	\\
25	&	8.22	&	0.104	&	0.0163	&	0.0153	\\
30	&	13.6	&	0.118	&	0.0161	&	0.0151	\\
40	&	30.9	&	0.153	&	0.0163	&	0.0153	\\
50	&	58.6	&	0.196	&	0.0168	&	0.0156	\\
100	&	452	&	0.339	&	0.0163	&	0.0157	\\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for morphological gradient transformation by boxes using several algorithm implementations applied to the $256 \times 256$ cthead image.\label{tab:perfGradient}}
\end{table}

\begin{table}[phtb]
\centering
\begin{tabular}{ccccc}
\hline
Radius	&	Basic	&	Histogram	&	Anchor	&	van Herk / Gil Werman	\\
\hline
1	&	0.0105	&	0.0208	&	0.0167	&	0.0146	\\
2	&	0.0217	&	0.0283	&	0.0157	&	0.0148	\\
3	&	0.0424	&	0.0352	&	0.0145	&	0.0154	\\
4	&	0.0752	&	0.0444	&	0.0142	&	0.0159	\\
5	&	0.117	&	0.049	&	0.0143	&	0.0159	\\
6	&	0.172	&	0.0573	&	0.014	&	0.0162	\\
7	&	0.24	&	0.0676	&	0.0139	&	0.0165	\\
8	&	0.324	&	0.072	&	0.0141	&	0.0169	\\
9	&	0.425	&	0.0804	&	0.0141	&	0.0175	\\
10	&	0.546	&	0.093	&	0.0143	&	0.0176	\\
15	&	1.5	&	0.14	&	0.0154	&	0.019	\\
20	&	3.19	&	0.178	&	0.0155	&	0.0204	\\
25	&	5.9	&	0.235	&	0.0164	&	0.0219	\\
30	&	9.9	&	0.274	&	0.0175	&	0.0236	\\
40	&	23	&	0.383	&	0.0193	&	0.0272	\\
50	&	45.3	&	0.522	&	0.0213	&	0.0298	\\
100	&	421	&	1.6	&	0.0346&		0.0493	\\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for morphological opening by boxes using several algorithm implementations applied to the $256 \times 256$ cthead image.\label{tab:perfOpening}}
\end{table}

\begin{table}[phtb]
\centering
\begin{tabular}{ccccc}
\hline
Threads	&	Basic	&	Histogram	&	Anchor	&	van Herk / Gil Werman	\\
\hline
1	&	3.37	&	2.34	&	1.23	&	1.02	\\
2	&	1.69	&	1.64	&	0.73	&	0.615	\\
3	&	1.14	&	1.15	&	0.46	&	0.367	\\
4	&	0.871	&	0.866	&	0.39	&	0.312	\\
5	&	1.06	&	1.03	&	0.505	&	0.437	\\
6	&	1.01	&	0.949	&	0.489	&	0.422	\\
7	&	0.963	&	0.866	&	0.493	&	0.422	\\
8	&	0.918	&	0.849	&	0.506	&	0.453	\\
9	&	0.947	&	0.897	&	0.548	&	0.491	\\
10	&	0.92	&	0.874	&	0.568	&	0.506	\\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for morphological opening by boxes using several algorithm implementations applied to the $3200 \times 2400$ image.\label{tab:perfThreads}}
\end{table}

\begin{table}[phtb]
\centering
\begin{tabular}{cccc}
\hline
Radius	&	Histogram	&	van Herk / Gil Werma	&	Anchorn	\\
\hline
1	&	0.354	&	0.379	&	0.532	\\
2	&	0.486	&	0.329	&	0.455	\\
3	&	0.62	&	0.338	&	0.411	\\
4	&	0.74	&	0.763	&	0.995	\\
5	&	0.862	&	0.757	&	0.95	\\
6	&	1.03	&	0.759	&	0.968	\\
7	&	1.17	&	0.77	&	0.936	\\
8	&	1.31	&	0.768	&	0.894	\\
9	&	1.52	&	1.11	&	1.28	\\
10	&	1.67	&	1.08	&	1.31	\\
15	&	2.42	&	1.09	&	1.25	\\
20	&	3.13	&	1.16	&	1.29	\\
25	&	3.84	&	1.14	&	1.26	\\
30	&	4.64	&	1.17	&	1.3	\\
40	&	5.95	&	1.19	&	1.23	\\
50	&	7.6	&	1.22	&	1.41	\\
100	&	19	&	1.36	&	1.34	\\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for morphological opening by boxes using several algorithm implementations applied to the $256 \times 256$ cthead image.\label{tab:perf2Dpoly}}
\end{table}

\begin{table}[phtb]
\centering
\begin{tabular}{ccccc}
\hline
Radius	&	Direct	&	Moving window	&	Separable direct	&	Separable moving window	\\
\hline
1	&	0.00355	&	0.00498	&	0.00487	&	0.00668	\\
2	&	0.00642	&	0.00706	&	0.0054	&	0.00679	\\
3	&	0.0127	&	0.00761	&	0.00861	&	0.00756	\\
5	&	0.0259	&	0.00936	&	0.00803	&	0.00749	\\
10	&	0.121	&	0.0162	&	0.0129	&	0.00765	\\
15	&	0.35	&	0.0224	&	0.0206	&	0.00755	\\
20	&	0.771	&	0.0252	&	0.0278	&	0.0076	\\
40	&	5.59	&	0.059	&	0.0699	&	0.00871	\\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for mean filters using direct and sliding window implementations applied to the $256 \times 256$ cthead image.\label{tab:perfRankMean}}
\end{table}

\begin{table}[phtb]
\centering
\begin{tabular}{ccccc}
\hline
Radius	&	Direct	&	Huang	&	Separable direct	&	Separable huang	\\
\hline
1	&	0.00642	&	0.00758	&	0.0071	&	0.0093	\\
5	&	0.0448	&	0.0186	&	0.0153	&	0.00913	\\
10	&	0.177	&	0.0265	&	0.0291	&	0.00895	\\
15	&	0.445	&	0.0342	&	0.0362	&	0.00876	\\
20	&	0.93	&	0.043	&	0.0444	&	0.00925	\\
40	&	6.21	&	0.061	&	0.0899	&	0.00936	\\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for median filters using direct and sliding window implementations applied to the $256 \times 256$ cthead image. The separable median is an approximation and therefore doesn't produce the same result as the direct or sliding algorithms.\label{tab:perfRankMedian}}
\end{table}

\begin{table}[phtb]
\centering
\begin{tabular}{ccccc}
\hline
Size  &  Direct Mean &  Box Mean & Direct sigma & Box sigma\\
1	&	0.557	&	0.916	&	0.858	&	1.45	 \\
2	&	1.15	&	0.924	&	1.64	&	1.56	 \\
3	&	2.08	&	0.932	&	2.74	&	1.56	 \\
4	&	3.47	&	0.92	&	4.34	&	1.55	 \\
5	&	5.13	&	0.935	&	6.34	&	1.55	 \\
6	&	7.06	&	1.08	&	8.53	&	1.62	 \\
7	&	9.42	&	1.08	&	11.3	&	1.63	 \\
8	&	12.2	&	0.922	&	14.5	&	1.63	 \\
9	&	15.3	&	0.924	&	18	&	1.63	 \\
10	&	18.9	&	0.925	&	22.1	&	1.63	 \\
15	&	43.3	&	0.929	&	50.2	&	1.64	 \\
20	&	80.1	&	0.936	&	91.8	&	1.64	 \\
25	&	131	&	0.939	&	149	&	1.64	 \\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for mean and standard deviation using direct and box convolution implementations applied to the $3200 \times 2400$ image.\label{tab:perfBoxConv}}
\end{table}

\begin{table}[phtb]
\centering
\begin{tabular}{ccccc}
\hline
Size  &  Direct Mean &  Box Mean & Direct sigma & Box sigma\\
\hline
1	&	2.1	&	0.942	&	2.75	&	1.57	\\
2	&	1.06	&	0.466	&	1.38	&	0.805	\\
3	&	0.716	&	0.391	&	0.932	&	0.551	\\
4	&	0.538	&	0.335	&	0.846	&	0.453	\\
5	&	0.659	&	0.327	&	0.902	&	0.525	\\
6	&	0.616	&	0.298	&	0.728	&	0.47	\\
7	&	0.579	&	0.285	&	0.758	&	0.466	\\
8	&	0.558	&	0.27	&	0.729	&	0.473	\\
9	&	0.614	&	0.289	&	0.771	&	0.467	\\
10	&	0.578	&	0.281	&	0.732	&	0.476	\\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for mean and standard deviation using direct and box convolution implementations applied to the $3200 \times 2400$ image.\label{tab:perfBoxConv-threads}}
\end{table}
